---
title: "evolSOM vignette"
author: "Santiago Prochetto"
date: Last modified "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About evolSOM

# About

In recent years, we have witnessed a surge in technological advancements in the field of life sciences. Researchers now have access to extensive biobanks and can collect vast amounts of multimodal data, including data from omics studies across multiple organisms, which is crucial for understanding evolution. As technical developments continue to progress, the demand for information systems capable of managing heterogeneous data is on the rise. One of the key challenges of this new era is the development of robust tools capable of integrating diverse data types to uncover hidden patterns and relationships among heterogeneous sources. We here introduce evolSOM, a novel tool based on unsupervised machine learning that facilitates the discovery of variations between multimodal data and offers an intuitive way to visualize hidden patterns within complex data matrices. Specifically, evolSOM clusters genes and/or phenotypic traits based on their expression patterns and maps these biological variables across different species to assess their conservation throughout evolution. This vignette provides a detailed overview of the most essential functions within the evolSOM package and outlines the overall process for analyzing the displacement of biological variables. The displacement of biological variables is a crucial aspect, as it allows us to uncover intrinsic evolutionary processes and changes. The general steps for using evolSOM are as follows:

1.  Import data and perform scaling.

2.  Construct a reference SOM (Self-Organizing Map).

3.  Map biological variables within the reference SOM.

4.  Characterize displacement types.

5.  Create a network dataframe to quantify and visualize displacements.

6.  Generate displacement plots to gain insights into the data.

# Installation

```{r setup, warning=FALSE}
devtools::install_github("sanprochetto/evolSOM")
library(evolSOM)
library(ggplot2)
```

# Example: Using evolSOM to Discover Potential Gene Drivers for Leaf Development in Grasses

In this example, we will demonstrate how to use evolSOM to identify potential gene drivers for leaf development in grasses. We'll go through the following steps:

## Step 1: Importing Data and Scaling

### 1.1 Data Import

First, we need to import our dataset. The dataset consists of a dataframe containing a list of genes and phenotypic traits. These genes and traits have been measured in leaf segments at four different developmental stages and across three closely related species. Additionally, we require an extra dataframe providing useful information about each biological variable. This information might include details such as the variable type (gene or phenotypic trait), the biological processes in which the variable is involved, the gene family to which the variable belongs, and more.

```{r}
#Traits expression data (genes + phenotype)
datSOM <- datSOM
head(datSOM)

#Traits names and information (tags) 
classSOM <- classSOM #First column must be name of biological variable/gen
```

### 1.2 Scaling

After importing the data, the next crucial step is to scale each biological variable to unit variance. This ensures that all variables have the same level of influence during the analysis, preventing any variable from dominating the results due to differences in scale. Scaling is essential for accurate and meaningful insights from evolSOM.

We can achieve scaling using the scale_species() function, which scales the data from each species individually. The output is a list containing one dataframe for each species. The reference species should be the first element in the list:

```{r}
datSOM.scaled <- scale_condition(datSOM[,9:12], datSOM[,5:8], datSOM[,1:4], 
                               rnames= row.names(datSOM),
                               condition_names = c("C3", "PK", "C4"))
head(datSOM.scaled[[1]])

head(datSOM.scaled[[2]])

head(datSOM.scaled[[3]])

```

## Step 2 - Build a reference SOM

### 2.1 Optimal size of the grid

Before constructing the reference SOM, it's essential to estimate the optimal size of the grid. The optimal size is where all expression patterns in your data are accurately represented by neurons, ensuring that each pattern is unique. The grid size depends on the number of conditions under which the data was measured. Estimating the grid size manually can be challenging, time-consuming, and may require several iterations, especially for those who are not familiar with Self-Organizing Maps (SOM). To streamline this process, we provide a script that automates the search for the optimum grid size using the opt_map_size() function. The result, map_size, will provide you with the optimal grid dimensions.

```{r}
map_size <- opt_map_size(data = na.omit(datSOM.scaled[[1]]), # Reference species data (without NA)
                     threshold = 0.8,          # Maximum allowed correlation for neurons
                     init_dim = 5,  # Initial grid size (e.g., 5x5)
                     max_iterations = 10)   # Maximum iterations when training the SOM model 

map_size
```

### 2.2 Build reference SOM

To build the reference SOM, we use functions from the kohonen and aweSOM packages:

```{r}
train.data <- as.matrix(na.omit(datSOM.scaled[[1]])) # Select variables for the reference condition
dim1<- map_size[["grid_dim1"]]
dim2<- map_size[["grid_dim2"]]
set.seed(100) # Set the random number generator seed for reproducibility
# Initialize the SOM using the aweSOM package
  init <- aweSOM::somInit(train.data, dim1, dim2)
  # Train the SOM using the kohonen package
  model.som <- kohonen::som(train.data,
                          grid = kohonen::somgrid(dim1, dim2, "hexagonal"), 
                    rlen = 1000, alpha = c(0.1, 0.001), 
                    dist.fcts = "euclidean", init = init)
  
  # Visualize the reference SOM using the aweSOM package
  aweSOM::aweSOMplot(som = model.som, type = "Line", data = train.data, 
           variables = colnames(train.data), 
           values = "prototypes", size= 400)
```

## Step 3 - Map biological variables within the reference SOM.

### 3.1 Create mappings

Let's briefly explain how to map biological variables within the reference SOM. In this step, we utilize the create_mappings() function to create mappings and class assignments based on a trained model and input data. The mappings object stores valuable information about the mappings, such as class assignments and number of biological variables allocated to each neuron.

```{r}
mappings <- create_mappings(model = model.som,
                            train_data = train.data,
                            data_list = datSOM.scaled, 
                            condition_names = names(datSOM.scaled), 
                            classes = classSOM)

#Explore info inside mappings:
head(mappings[["classes"]], 15)
head(mappings[["neurons_sizes"]],15)
```

### 3.2 Neurons sizes graph

Now, let's take a look at the neuron sizes graph to gain insights into the number of biological variables residing within each neuron:

```{r,fig.align = 'center', fig.width=5, fig.height=5, fig.retina= 3}

figure <- ggplot(mappings[["neurons_sizes"]], aes(x = Neuron, y = size, fill = condition)) +
   geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
    scale_x_continuous(breaks = seq(1, 6, 1)) +
  scale_fill_manual(values = c("#7ed977", "#008f77", "#009ca6" ))+
 theme_bw() 

figure
```

## Step 4 - Detect and count displacements

### 4.1 - Characterizing displacement types

To uncover hidden evolutionary patterns, the package identifies movements, or displacements, of biological variables among neurons within the Self-Organizing Map (SOM). When a biological variable from a test species is mapped to the same neuron as in the control condition, it is considered preserved. If it is mapped to a different neuron, it is considered displaced. These biological variable movements can also be categorized into various types, such as early, delayed, flip, and more.

You can automatically detect displacement types using the create_displacements() function:

```{r}
displacements<-  detect_displacements(model.som,
                                      delay_threshold = 0.75,
                                      flip_threshold = -0.85)

displacements
```

The displacements object will provide information about the detected displacement types. If you need to further characterize a displacement based on additional criteria, you can manually edit the displacements dataframe as needed. This step allows for a more detailed exploration of displacement patterns in your data.

### 4.2 - Creating a Network Dataframe to Count and Visualize Displacements

To gain insights into the nature of the displacements, we employ the net_edges() function to generate a displacement graph. This function constructs a dataframe that counts the number of displacements between neurons and condition pairs

```{r}
net_edges_C3toPK <- net_edges(mappings =  mappings, #object returned from mapping()
                                     "Reference.sp", "PK",  #Condition names: "start", "end"
                                     add_disp_type = TRUE,  # choose to add displacement types
                                     disp_type = displacements) #where to find disp types info
head(net_edges_C3toPK, 10)
```

The net_edges_C3toPK dataframe provides valuable information about the number and types of displacements between neurons and conditions. You can use this data to create visualizations and gain insights into the changes and relationships between the reference species (C3) and the test species (PK)

## Step 5 - Plot the displacements

### 5.1 General displacements

You can use the plot_displacements() function to observe the extent of displacements of biological variables between species. With this visualization, you can gain insights into the extent and quality of displacements between condition/species.

```{r,fig.align = 'center', fig.width=8, fig.height=5, fig.retina= 3}
 figure <- plot_displacements(mappings = mappings,  # Looks for nodes inside mapping mapping[["neurons_sizes"]]
                 links =  net_edges_C3toPK, #the arrows in the graph
                 reference_condition = "C3", #used for determine the size of the neurons
                 layout="circle",  #from igraph: circle, linear, grid, etc.
                 node_order=c(1,4,5,2,6,3),  #use a specific order for visualization purposes
                 color_edges = "disp_type",  #Color the arrows by...
                 color_scale=c("#ef466e","#1189b1ff","#ffd166ff", "#a1a5afff", "black"))
                 
figure
                 
```

### 5.2 Displacements of a specific group of biological variables

You can also select a subset of biological variables and visualize their displacements. For instance, based on the work of Prochetto et al., 2023, we have compiled a list of 25 genes related to the "photorespiration" process. This allows us to investigate whether there's a distinct pattern of displacement for this particular group of genes.

```{r, fig.align = 'center',  fig.width=8, fig.height=5, fig.retina= 3}
# Photorespiration (25) gene list
photo.genes <- c("OG0000295", "OG0000378", "OG0000783", "OG0000964", "OG0001962",
               "OG0002384", "OG0003519", "OG0003992", "OG0004041", "OG0004061",
               "OG0004294", "OG0004978", "OG0005385", "OG0005556", "OG0006771",
               "OG0007574", "OG0007822", "OG0008057", "OG0010248", "OG0012065",
               "OG0012541", "OG0000516", "OG0011137", "OG0001049", "OG0005515")

# Select genes from mappings[["classes"]]
selected_variables <- mappings[["classes"]][mappings[["classes"]]$bio_var %in% photo.genes,]
mappings_selected_variables <- mappings
mappings_selected_variables$classes <- selected_variables



# Build net_edges for the list using the net_edges function
net_edges_pk_photo <- net_edges(mappings_selected_variables, 
                                     "C3", "PK",
                                     add_disp_type = TRUE,
                                     disp_type = displacements)
head(net_edges_pk_photo, 10)

plot_displacements(mappings = mappings_selected_variables,
                links = net_edges_pk_photo,
                reference_condition = "C3",
                layout="circle", 
                node_order = c(1,4,5,2,6,3),
                color_edges = "disp_type",
                color_scale=c("#ef466e","#1189b1ff","#ffd166ff", "#a1a5afff", "black")
                )
```

With this analysis, you can explore the displacement patterns for this specific group of genes associated with "photorespiration." This can provide valuable insights into how these genes expression pattern have changed and their relationships between the reference species (C3) and the test species (PK).

### 5.3 Photosynthesis as a Case Study

  In previous steps we integrated transcriptomic data with anatomical information from four leaf segments representing different developmental stages. After scaling the data, an optimal map size of 2x3 was determined for our dataset comprising 9757 biological variables. The resulting map consisted of 6 neurons, each exhibiting distinct expression patterns. Following this, we constructed a reference SOM utilizing data from the C3 species, and subsequently, we mapped the data from PK and C4 onto this reference SOM for further analysis.
  Our analysis centered on inner bundle sheath (IBS) cell size, a crucial phenotypic trait in C4 anatomy. Our results show that the  phenotypc trait IBS cell area is preserved between C3 and PK species (In both species was mapped to neuron 1) but it is displaced to neuron 6 in C4 species. We can acces to this information looking for the "classes" dataframe inside "mappings" object created in step 3.1.
  

```{r}
head(mappings[["classes"]], 10)

```

Now we can select all biological variables mapped to neuron 1 in both C3 and PK
```{r}
#
selected_variables <- mappings[["classes"]][mappings$classes$C3 %in% 1 & mappings$classes$PK %in% 1 ,]

head(selected_variables) # 404 biological variables
```

We also can plot these displacement regarding from C3 and PK to C4
```{r,fig.align = 'center', fig.width=8, fig.height=5, fig.retina= 3}
# Select genes from mappings[["classes"]]
mappings_selected_variables <- mappings
mappings_selected_variables$classes <- selected_variables

# Build net_edges for the list using the net_edges function
net_edges_selected_variables <- net_edges(mappings_selected_variables, 
                                     "C3", "C4",
                                     add_disp_type = TRUE,
                                     disp_type = displacements)

head(net_edges_selected_variables, 10)

plot_displacements(mappings = mappings_selected_variables,  # Looks for nodes inside mapping mapping[["neurons_sizes"]]
                 links =  net_edges_selected_variables, #the arrows in the graph
                 reference_condition = "C3", #used for determine the size of the neurons
                 layout="circle",  #from igraph: circle, linear, grid, etc.
                 node_order=c(1,4,5,2,6,3),  #use a specific order for visualization purposes
                 color_edges = "disp_type",  #Color the arrows by...
                 color_scale=c("#ef466e","#1189b1ff","#ffd166ff", "#a1a5afff", "black"))
```

Select biological variables displaced together with IBS Cell size:
```{r}
head(selected_variables)  #IBS cell size is mapped to Neuron 6 in C4.

ibs_displaced_bv <- selected_variables[selected_variables$C4 %in% 6,1]
ibs_displaced_bv

```

Among the displaced genes were two known components of suberin biosynthesis, a characteristic polymer in C4 bundle sheath cell walls. Additionally, our analysis identified two displaced transcription factors, suggesting potential regulatory mechanisms underlying the observed phenotypic shift.
We can also use this list of biological variables to perform some kind of enrichment analysis, depending on the annotation information available. For example a GO enrichment analysis showed that this group of genes is significantly enriched in suberin biosynthetic process, pyruvate transport and cell wall modification, among others (Prochetto et al., 2024).


## Citation

For more information, please refer to our paper:

"insert reference here"
